### 2014-05-13 Paper1/paper1.tex
Updated paper, figs

\section*{Introduction}
As machines interact more and more with humans, there is a growing need for a rigorous way to understand what people are trying to do, even if they become disturbed, distracted or influenced by pathology. One can assume that a person who is familiar with a particular task will try to do the same thing each time provided all other conditions are equal. This presumes that what people do is on average what they intended to do and that their goal remains unchanged. However, if there is a disturbance, it is unclear what happens to a person’s intentions because action is decoupled from intention. Lacking a means to discern that intention and changes in intention leads to challenges in human machine interactions. Hence a comprehensive method for understanding what someone means to do (rather than what they actually do or claim to have wanted to do) as they interact within uncertain environments will prove useful in any area where a person might be thwarted from their intent such as national security, operation of machinery, athletic and music performance, rehabilitation, human augmentation, and artificial intelligence.

@@

Attempts to deduce motor intent in the past have focused on the assumed spring-like properties of human muscles. Springs produce a force according to their impedance and stretch. By measuring force, impedance, and position, Gomi and Kawato\cite{gomi1997human} were able to deduce stretch and thereby infer the muscle’s equilibrium point. Supporters of the so-called lambda model\cite{feldman1995origin} hypothesized that this equilibrium point represented the intent of a movement. Upon inspection of the equilibrium point as it evolved in time, it was clear that it was highly complex and often not anatomically realizable. Therefore, it could not well-represent the intent of a simple reaching movement. We propose a revised method---intent extraction---that deduces intended trajectory, even in the face of disturbances. This method describes a class of filters that map a disturbance, process, actuator, and realized trajectory to an intended trajectory. In human reaching, the process (a dynamic model of the arm), the actuator (viscoelastic muscles), and the realized trajectory are all readily accessible via measurement and modeling. The underlying framework should be extensible to whole-body movement or even more abstract, psychological processes.

Collectively, the general problem of being able to best extract someone’s intent is critical to many fields, and hinges on our understanding of the underlying mechanisms through which control is accomplished. We have excellent models of the mechanical properties of arm and muscle because the physics that describe their gross behavior is so well-understood and the need to determine particular physical parameter values has long been clear. This wealth of previous work provides strong footing for our technique, which in turn should yield valuable information not previously available by other means. In fields where the process and actuators are not yet well-understood, this approach may help to structure the search for them.

Understanding the intent has many benefits, and one may be a better way to understand the composition of discrete movements into their subunits. Recent research has strived to understand the composition of movements intention during both infant development\cite{von1979observations} and stroke recovery\cite{rohrer2004submovements}, distinct straight-line subunits with bell-shaped speed profiles increasingly overlap and blend until, as in healthy adult reaching\cite{woodworth1899accuracy}, they become indistinguishable. Flash and Henis \cite{flash1991arm} demonstrated that despite this apparent blending, cued corrections can only be produced at predictable places within the reach. They also showed that these modifications appear to take place by superimposing additional straight-line subunits onto the original movement. Novak et al. \cite{novak2002use} detected corrective subunits following disturbance during knob turning. These studies show planning and correction in discrete subunits, but did not analyze the actual intent and hence cannot be compared to a known, correct trajectory except in the trivial case that movement is undisturbed. A disturbed trajectory’s intention is hidden without proper inversion of the model plan for movement. Here we develop a method to extract intent and gain a clearer understanding of subunits from disturbed motions.

@@

In fact, this approach can be generalized to any system with an intention that governs the control law, and hence is agnostic to the form of controller. In order to prove this, we introduce one additional layer of complexity by separating the plant into its process and actuator components. Modeling the actuator explicitly allows us to compare our $x_d$ against lambda models \cite{feldman1995origin} and their hypothetical equilibrium trajectory, $x_e$. Additionally, this enables a more general derivation. Without loss of generality, we describe both as instantaneously linear differential equations in some space, $x$:
\begin{equation}
\underbrace{\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+\overbrace{\sum_{m=1}^M A_m(x-x_e)^{(m)}}^\text{Actuator}}_\text{Plant}+E=0
\end{equation}
Controlling the process as intended requires a specific relationship between the internal model, choice of actuator equilibria $x_e$, and impedances $A_n$:
\begin{equation}
\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m(x_d-x_e)^{(m)}}^\text{Actuator}=0
\end{equation}
Because the actuator's reference trajectory is shared between both representations, we can solve both equations for $\sum_{m=1}^M A_mx_e^{(m)}$, set them equal, and collect the remaining actuator terms into a single feedback component:
\begin{equation}
\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+E=\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m (x_d-x)^{(m)}}^\text{Feedback (Actuator)}
\end{equation}
We now recover our familiar model and learn its relationship with $x_e$. While a proper choice of $x_e$ was critical in controlling the actuator, a complete model relating actual and intended trajectory does not and need not reflect it. The actuator's equilibrium is not the process's equilibrium unless $x_d$ and all of its derivatives are zero. This upholds Gomi and Kawato's finding that the arm muscles' equilibrium does represent reaching intent\cite{gomi1997human}.

This general form reveals the conditions needed to solve for the intention of an arbitrary process. We can solve for $x_d$ if the actuator's instantaneous impedance is known and either the plant or its highest order term can be inverted. Moreover, the plant model does not need to be explicitly formulated. In cases like national security or athletic performance, the plant (a bomb in a crowded square or an opposing team) may be too complex to model. By substituting a lookup table made up of past state-outcome pairs for an explicit process model, it becomes possible to determine desired outcome even when the state is disturbed. It is not necessarily to allow a bomb to be planted or an opposing team to carry out a play in order to discover their plans.

\textcolor{red}{\textbf{This last bit needs work. Solving for the Feldman equilibrium addresses your prior objections and keeps everything in familiar forms throughout. I'm not sure if take-home points for the reader go here or in the discussion. In particular, this gets us back to skewering Feldman.}}

\subsection*{Generation of synthetic data using simple dynamic simulation}
We choose one prevailing arm model, that of Burdet et al. (2006), along with the anatomical mass distribution and geometry parameters of Winters (1990) to create a model for estimating intent. The model of Burdet et al (2006) was adapted to reflect measured physical parameters of the subject who provided the reach. Segment lengths were measured directly while gross body mass was self-reported. These values were converted to segment masses, centers of gravity, and moments of inertia inertia using anatomical landmarks and values from Dempster (1955) and Winters (1990). The muscle torque used to calculate stiffness was approximated by looking backward in time a small amount (500 microseconds). Desired trajectory in time was simulated as a minimum jerk, 5th order polynomial of duration 700ms starting and ending with zero velocity and acceleration.
Simulated reaches were 15 cm long, beginning 38 cm out from and 5.7 cm left of the right shoulder then proceeding right.  We used the following types of perturbing forces for each combination of distance and direction: 
Pulse forces applied in one of the the two directions perpendicular to the direction of movement began when the subject had moved either 10\% or 50\% of the distance to the target and lasted for 150ms.
Noise forces began once the subject moved 3 mm, and lasted for the duration of the motion. The forces were drawn from a white noise generator drawn at 1000 Hz with flat power spectral density of 1N, and then passed through a 4th order low-pass Butterworth filter with cutoff $10\pi$ rad/s.

@@

\subsection*{Indices of variance-based sensitivity}
Sobol-distributed matrices were generated using MATLAB’s sobolset() function and converted to parameter distributions (see Table 1) using inverse cumulative probability density. Distributions were clamped to the range of $\pm3$ standard deviations in order to avoid support for implausible parameter values. Combinations and calculations were made as prescribed by Saltelli et al. (2010) to arrive at direct and indirect sensitivity indices.

\subsection*{Human Subjects}
The human data trajectories analyzed here are drawn from a four subjects who gave informed consent in accordance with Northwestern University Institutional Review Board. Two male and two female right-handed subjects (ages 24 to 30) performed the reaches with their right arm and was not compensated. Subjects’ arm segment lengths were directly measured in situ while body mass was self-reported.

\subsection*{Apparatus}
A planar manipulandum (described in Patton and Mussa-Ivaldi, 2004) was programmed to compensate and minimize any handle friction or mass. The MATLAB XPC-TARGET package (Natick, MA) was used to render this force environment at 1000 Hz and data were collected at 1000 Hz.  Visual feedback was performed at 60 Hz using OpenGL. Closed loop data transmission time (position measurement to completed rendering to recognition of rendering by the position measurement system) was <8 ms, ensuring a visual delay less than one 60 Hz frame.

@@

\subsection*{Model Viability}
Sensitivity analysis fortunately revealed that variance in intended trajectories due to estimated uncertainty in model parameters is lower than the natural variation in undisturbed motions.  Simulated point-to-point reaches disturbed by either filtered white noise forces or a pulse force perpendicular to the direction of the reach (figure 1) were extracted in the presence of 220,000 variations upon the parameters (table 1) according to the methods of Saltelli et al. (2011). Expected variation due to parameter uncertainty was 0.86 microns for pulse forces and 1.5 microns for filtered white forces. Variation in recorded undisturbed point-to-point reaching under the same time and reach distance conditions was 2.93 microns. These variations both fall beneath the instantaneous accuracy of the robot recording the motion (tenths of a millimeter). In other words, while some inversion processes might be highly sensitive to model uncertainty, our process for recovering intent from action is much more precise than its input.

@@

@@

>>>>

\section*{Introduction}
As machines interact more and more with humans, there is a growing need for a rigorous way to understand what people are trying to do, even if they become disturbed, distracted or influenced by pathology. One can assume that a person who is familiar with a particular task will try to do the same thing each time provided all other conditions are equal. This presumes that what people do is on average what they intended to do and that their goal remains unchanged. However, if there is a disturbance, it is unclear what happens to a person's intentions because action is decoupled from intention. Lacking a means to discern that intention and changes in intention leads to challenges in human machine interactions. Hence a comprehensive method for understanding what someone means to do (rather than what they actually do or claim to have wanted to do) as they interact within uncertain environments will prove useful in any area where a person might be thwarted from their intent such as national security, operation of machinery, athletic and music performance, rehabilitation, human augmentation, and artificial intelligence.

@@

Attempts to deduce motor intent in the past have focused on the assumed spring-like properties of human muscles. Springs produce a force according to their impedance and stretch. By measuring force, impedance, and position, Gomi and Kawato\cite{gomi1997human} were able to deduce stretch and thereby infer the muscle's equilibrium point. Supporters of the so-called $\lambda$ model\cite{feldman1995origin} hypothesized that this equilibrium point represented the intent of a movement. Upon inspection of the equilibrium point as it evolved in time, it was clear that it was highly complex and often not anatomically realizable. Therefore, it could not well-represent the intent of a simple reaching movement. We propose a revised method---intent extraction---that deduces intended trajectory, even in the face of disturbances. This method describes a class of filters that map a disturbance, process, actuator, and realized trajectory to an intended trajectory. In human reaching, the process (a dynamic model of the arm), the actuator (viscoelastic muscles), and the realized trajectory are all readily accessible via measurement and modeling. The underlying framework should be extensible to whole-body movement or even more abstract, psychological processes.

Collectively, the general problem of being able to best extract someone's intent is critical to many fields, and hinges on our understanding of the underlying mechanisms through which control is accomplished. We have excellent models of the mechanical properties of arm and muscle because the physics that describe their gross behavior is so well-understood and the need to determine particular physical parameter values has long been clear. This wealth of previous work provides strong footing for our technique, which in turn should yield valuable information not previously available by other means. In fields where the process and actuators are not yet well-understood, this approach may help to structure the search for them.

Understanding the intent has many benefits, and one may be a better way to understand the composition of discrete movements into their subunits. Recent research has strived to understand the composition of movements intention during both infant development\cite{von1979observations} and stroke recovery\cite{rohrer2004submovements}, distinct straight-line subunits with bell-shaped speed profiles increasingly overlap and blend until, as in healthy adult reaching\cite{woodworth1899accuracy}, they become indistinguishable. Flash and Henis \cite{flash1991arm} demonstrated that despite this apparent blending, cued corrections can only be produced at predictable places within the reach. They also showed that these modifications appear to take place by superimposing additional straight-line subunits onto the original movement. Novak et al. \cite{novak2002use} detected corrective subunits following disturbance during knob turning. These studies show planning and correction in discrete subunits, but did not analyze the actual intent and hence cannot be compared to a known, correct trajectory except in the trivial case that movement is undisturbed. A disturbed trajectory's intention is hidden without proper inversion of the model plan for movement. Here we develop a method to extract intent and gain a clearer understanding of subunits from disturbed motions.

@@

It is important to distinguish our intended trajectory, $x_d$, from another variable often considered, the equilibrium point of the muscle actuators. Importantly, this has been claimed to be the intended movement. To consider this, we also generalize our approach to any system with a moving intention that governs the control law. This introduces two additional layers of complexity by separating the plant (arm) into its process (dynamics) and actuator components (muscles) and considering the actuator's equilibrium, $x_e$ (sometimes represented as $\lambda$). The feedforward component must use and share this actuator and hence use the actuator's equilibrium. Without loss of generality, we represent the process $P$ and actuator $A$ as linear (or linearizable) operations in some space, $x$:
\begin{equation}
\underbrace{\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+\overbrace{\sum_{m=1}^M A_m(x-x_e)^{(m)}}^\text{Actuator}}_\text{Plant}+E=0
\end{equation}
The internal model of the process ($\hat{P}$ and $\hat{E}$) predicts system actions, external disturbances, and impedance responses in order to determine $x_e$ that $x$ will track a desired path $x_d$.    
\begin{equation}
\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m(x_d-x_e)^{(m)}}^\text{Actuator}=0
\end{equation}
We solve this for $x_e$ and substituting into eq 4 gives:
\begin{equation}
\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+E=\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m (x_d-x)^{(m)}}^\text{Feedback (Actuator)}
\end{equation}

Note $x_e$ vanishes, recovering our familiar model. While a proper choice of $x_e$ is critical to control, our approach does not require a model relating equilibrium and intended trajectory. Note also that the actuator's equilibrium $x_e$ is not the process's equilibrium unless all derivatives of $x_d$ are zero. This upholds Gomi and Kawato's finding that the arm muscles' equilibrium does not represent reaching intent\cite{gomi1997human}.

In summary, if the highest order coefficient of the process $\hat{P}_N$ can be inverted and the impedances can be modeled, it is possible to solve for $x_d^{(N)}$ and integrate for $x_d$ revealing the intended trajectory. This general form reveals the conditions needed to solve for the moving intent of an arbitrary dynamic process.

\subsection*{Generation of synthetic data using simple dynamic simulation}
We choose one prevailing arm model, that of Burdet et al. \cite{burdet2006stability}, along with the anatomical mass distribution and geometry parameters of Winter \cite{winter2009biomechanics} to create a model for estimating intent. The model of Burdet et al \cite{burdet2006stability} was adapted to reflect measured physical parameters of the subject who provided the reach. Segment lengths were measured directly while gross body mass was self-reported. These values were converted to segment masses, centers of gravity, and moments of inertia inertia using anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics}. The muscle torque used to calculate stiffness was approximated by looking backward in time a small amount (500 microseconds). Desired trajectory in time was simulated as a minimum jerk, 5th order polynomial of duration 700ms starting and ending with zero velocity and acceleration.
Simulated reaches were 15 cm long, beginning 38 cm out from and 5.7 cm left of the right shoulder then proceeding right.  We used the following types of perturbing forces for each combination of distance and direction: 
Pulse forces applied in one of the the two directions perpendicular to the direction of movement began when the subject had moved either 10\% or 50\% of the distance to the target and lasted for 150ms.
Noise forces began once the subject moved 3 mm, and lasted for the duration of the motion. The forces were drawn from a white noise generator drawn at 1000 Hz with flat power spectral density of 1N, and then passed through a 4th order low-pass Butterworth filter with cutoff $10\pi$ rad/s.

@@

\subsection*{Indices of variance-based sensitivity}
Sobol-distributed matrices were generated using MATLAB's sobolset() function and converted to parameter distributions (see Table~\ref{tab:parameters}) using inverse cumulative probability density. Distributions were clamped to the range of $\pm3$ standard deviations in order to avoid support for implausible parameter values. Combinations and calculations were made as prescribed by Saltelli et al. \cite{saltelli2010variance} to arrive at direct and indirect sensitivity indices.

\subsection*{Human Subjects}
The human data trajectories analyzed here are drawn from a four subjects who gave informed consent in accordance with Northwestern University Institutional Review Board. Two male and two female right-handed subjects (ages 24 to 30) performed the reaches with their right arm and was not compensated. Subjects' arm segment lengths were directly measured \textit{in situ} while body mass was self-reported.

\subsection*{Apparatus}
A planar manipulandum (described in Patton and Mussa-Ivaldi \cite{patton2004robot}) was programmed to compensate and minimize any handle friction or mass. The MATLAB XPC-TARGET package (Natick, MA) was used to render this force environment at 1000 Hz and data were collected at 1000 Hz.  Visual feedback was performed at 60 Hz using OpenGL. Closed loop data transmission time (position measurement to completed rendering to recognition of rendering by the position measurement system) was <8 ms, ensuring a visual delay less than one 60 Hz frame.

@@

\subsection*{Model Viability}
Sensitivity analysis fortunately revealed that variance in intended trajectories due to estimated uncertainty in model parameters is lower than the natural variation in undisturbed motions.  Simulated point-to-point reaches disturbed by either filtered white noise forces or a pulse force perpendicular to the direction of the reach (figure 1) were extracted in the presence of 220,000 variations upon the parameters (table 1) according to the methods of Saltelli et al. \cite{saltelli2010variance}. Expected variation due to parameter uncertainty was 0.86 microns for pulse forces and 1.5 microns for filtered white forces. Variation in recorded undisturbed point-to-point reaching under the same time and reach distance conditions was 2.93 microns. These variations both fall beneath the instantaneous accuracy of the robot recording the motion (tenths of a millimeter). In other words, while some inversion processes might be highly sensitive to model uncertainty, our process for recovering intent from action is much more precise than its input.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{fig1.eps}
\end{center}
\caption{
{\bf Simulated data illustrating tautology of extraction absent parameter error across pulse and filtered Gaussian noise disturbance types.} Intent is modeled as a minimum jerk, 5th order polynomial. Forces experienced are combined with intent via Burdet et al.'s \cite{burdet2006stability} model to produce the simulated arm trajectory. Extraction to recover intention from arm and force trajectory follows.
}
\label{fig:synthetic}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{fig2.eps}
\end{center}
\caption{
{\bf Model sensitivity testing (from the variance-based method of Saltelli et al.\cite{saltelli2010variance}) reveals that the model is mostly sensitive to stiffness.} Direct sensitivity, in purple, shows how modeling error changes with the parameter. Indirect sensitivity, in teal, shows how modeling error from other parameters is affected by changes in the parameter. Variation in individual tests exceeds expected variance by two orders of magnitude.
}
\label{fig:sensitivity}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{fig3.eps}
\end{center}
\caption{
{\bf Subjects' disturbed hand trajectories and extraction of intended trajectories from them reveal that both variance and invariance of the motor plan can occur even in response to very large disturbances.} The hand path, in black, deviates from the green baseline, as a force pulse, gray arrows, are applied. This subject's extracted desired trajectories, in red (baseline in faded red), do not significantly differ from baseline during force application, but show some reaction afterwards.
}
\label{fig:anecdotes}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{fig4.eps}
\end{center}
\caption{
{\bf Comparison of subjects' perpendicular errors across disturbed and undisturbed conditions reveals that intention is statistically indistinguishable from undisturbed reaching up until 100ms after the onset of movement.} Hand errors in the presence of disturbance (black dots, black band is median $\pm$ simultaneous 95\% confidence interval on the median) deviate almost immediately from undisturbed hand errors (green dots, band). Movement intention (red dots, band) deviates at approximately 100ms after the onset of disturbance. The hand rejoins the intention approximately 500ms after the onset of disturbance.
}
\label{fig:grouptrends}
\end{figure}

@@

\begin{table}[!ht]
\caption{
\bf{Arm Parameter Values and Sensitivities}}
\begin{tabular}{|c|c|c c|c c|}
\hline
Parameter Name &
Units &
Nominal &
SD &
$\frac{S}{\text{Total}}$ &
$\frac{ST}{\text{Total}}$ \\ \hline
Upper Arm Length ($L_1$) &
m &
0.33 &
0.01 &
0.07 &
0 \\
Forearm Length ($L_2$) &
m &
0.34 &
0.01 &
0.02 &
0 \\
Upper Arm Center of Mass Ratio ($\frac{L_{m1}}{L_1}$) &
1 &
0.436 &
0.0695 &
0.01 &
0 \\
Forearm Center of Mass Ratio ($\frac{L_{m2}}{L_2}$) &
1 &
0.682 &
0.0431 &
0.02 &
0 \\
Gross Body Mass ($m_g$) &
kg &
79.36 &
3.1 &
0.01 &
0 \\
Upper Arm Mass Ratio ($\frac{m_1}{m_g}$) &
1 &
0.028 &
0.0029 &
0.01 &
0 \\
Forearm Mass Ratio ($\frac{m_2}{m_g}$) &
1 &
0.022 &
0.0025 &
0.04 &
0 \\
Upper Arm Radius of Gyration Ratio &
1 &
0.322 &
0.0161 &
0.01 &
0 \\
Forearm Radius of Gyration Ratio &
1 &
0.468 &
0.0234 &
0.01 &
0 \\
Shoulder Parallel Coordinate &
m &
-0.057 &
0.02 &
0.13 &
0.01 \\
Shoulder Perpendicular Coordinate &
m &
0.88 &
0.02 &
0.05 &
0 \\
Force Sensor Miscalibration $x\text{-axis}$ &
N &
0 &
0.231 &
0.06 &
0 \\
Force Sensor Miscalibration $y\text{-axis}$ &
N &
0 &
0.1067 &
0.06 &
0.02 \\
Force Sensor Gaussian Noise SD $x\text{-axis}$ &
N &
0 &
0.1653 &
0.01 &
0 \\
Force Sensor Gaussian Noise SD $y\text{-axis}$ &
N &
0 &
0.2869 &
0.01 &
0 \\
Torque-Invariant Impedance Misestimation Ratio &
1 &
1 &
0.15 &
0.22 &
0.04 \\
Torque-Varying Impedance Misestimation Ratio &
1 &
1 &
0.15 &
0.11 &
0.01 \\
Damping-to-Stiffness Ratio ($k_d$) &
$\text{sec}^{-1}$ &
0.0833 &
0.0125 &
0.06 &
0 \\
Reflex Impedance Scale Factor &
1 &
0.02 &
0.003 &
0.03 &
0 \\
Reflex Damping to Stiffness Ratio ($g_d$) &
$\text{sec}^{-1}$ &
2 &
0.3 &
0.03 &
0 \\ \hline
\end{tabular}
\begin{flushleft}Synthetic model parameters and their associated mean (nominal) values and standard deviations (SD) used to determine the sensitivity indices of Saltelli et al. \cite{saltelli2010variance}. Also shown are the resulting sensitivity indices. Sensitivity indices less than one-thousandth of the total sensitivity were reported as zero. Total sensitivity was $1.8 \mu m$, showing very small average deviations when using this approach.
\end{flushleft}
\label{tab:parameters}
\end{table}
